Can Symmetric Cryptography Be Liberated from the Von Neumann Style?

### 2024-10-13 paper.tex
typo ?
typo

Moreover, ciphers are now, by and large, designed to be run
efficiently on software. The era of hardware-based cryptography is
long gone, even for embedded
platforms~\citep{buchanan:nist-lightweight-crypto}. To achieve
high-throughput, ciphers are typically designed so as to admit a
bitsliced implementation. Bitslicing enables a form of parallel
execution \emph{within} machine words. We thus gain access to a form
of CPU-level ``scale out'' to larger machine words, in particular
through single-instruction multiple-data (SIMD) instructions. Doubling
the size of machine words yields nearly twice the throughput, without
any effort. Evidences tend to suggest that cryptographers obtain
bitsliced designs through sheer intellectual might. It is in fact the
sole purpose of the traditional ``Section 5'' scene to exhibit a
bitsliced witness in all the gory details.

@@

What would it take to turn this art form into an engineering
principle? To answer, we make the following contributions:
%
\begin{itemize}
\item we identify the algebraic structure necessary to model the state
  of modern symmetric cryptography primitives
  (Section~\ref{sec:spec}). Doing so, we delineate a mathematical
  language of software circuits, rooted in the categorical notion of
  functor to account for data containers~;
\item we give a formal account of bitslicing as a data representation
  change supported by suitably polymorphic definitions
  (Section~\ref{sec:bitslicing}). As expected when it comes to
  switching between data representations, parametricity plays a key
  role to justify the equivalence of the resulting programs~;
\item we give a formal account of fixslicing as a whole-program
  transformation (Section~\ref{sec:fixslicing}). We show how a
  fixsliced implementation can be obtained through equational
  reasoning. To support such reasoning, we crucially rely on the
  algebraic structure set forth in Section~\ref{sec:spec}.
\end{itemize}

@@

Bitslicing is the cryptographer's jargon for such a representation
change. In the case of \Skinny{}, we observe that only \coqe!shiftrows_!
imposes a strict composition order of \coqe!Rows.T! followed by
\coqe!Cols.T!. Aside from that, we can commute \coqe!Slice8.T! out the
composition. We would thus have the type \coqe!Rows.T (Cols.T bool)!
representing 16 bits and fitting snugly in a 32 bits machine word. The
overall composition \coqe!Slice8.T (Rows.T (Cols.T bool))! would take
8 registers in total, imposing a very reasonable amount of register
pressure on the CPU.
%
In effect, to return back to the jargon of compiler designers, we
propose to turn a structure-of-arrays (SoA) into an
array-of-structures (AoS), only we are dealing with bit-level
quantities here.

>>>>

Moreover, ciphers are now, by and large, designed to be run
efficiently on software. The era of hardware-based cryptography is
long gone, even for embedded
platforms~\citep{buchanan:nist-lightweight-crypto}. To achieve
high-throughput, ciphers are typically designed so as to admit a
bitsliced implementation. Bitslicing enables a form of parallel
execution \emph{within} machine words. We thus gain access to a form
of CPU-level ``scale out'' to larger machine words, in particular
through single-instruction multiple-data (SIMD) instructions. Doubling
the size of machine words yields nearly twice the throughput, without
any effort. Evidence tends to suggest that cryptographers obtain
bitsliced designs through sheer intellectual might. It is in fact the
sole purpose of the traditional ``Section 5'' scene to exhibit a
bitsliced witness in all the gory details.

@@

What would it take to turn this art form into an engineering
principle? To answer, we make the following contributions:
%
\begin{itemize}
\item we identify the algebraic structure necessary to model the state
  of modern symmetric cryptographic primitives
  (Section~\ref{sec:spec}). Doing so, we delineate a mathematical
  language of software circuits, rooted in the categorical notion of
  functor to account for data containers~;
\item we give a formal account of bitslicing as a data representation
  change supported by suitably polymorphic definitions
  (Section~\ref{sec:bitslicing}). As expected when it comes to
  switching between data representations, parametricity plays a key
  role to justify the equivalence of the resulting programs~;
\item we give a formal account of fixslicing as a whole-program
  transformation (Section~\ref{sec:fixslicing}). We show how a
  fixsliced implementation can be obtained through equational
  reasoning. To support such reasoning, we crucially rely on the
  algebraic structure set forth in Section~\ref{sec:spec}.
\end{itemize}

@@

Bitslicing is the cryptographer's jargon for such a representation
change. In the case of \Skinny{}, we observe that only \coqe!shiftrows_!
imposes a strict composition order of \coqe!Rows.T! followed by
\coqe!Cols.T!. Aside from that, we can commute \coqe!Slice8.T! out of the
composition. We would thus have the type \coqe!Rows.T (Cols.T bool)!
representing 16 bits and fitting snugly in a 32 bits machine word. The
overall composition \coqe!Slice8.T (Rows.T (Cols.T bool))! would take
8 registers in total, imposing a very reasonable amount of register
pressure on the CPU.
%
In effect, to return back to the jargon of compiler designers, we
propose to turn a structure-of-arrays (SoA) into an
array-of-structures (AoS), only we are dealing with bit-level
quantities here.
